The Strucplot Framework:\\ Visualizing Multi-way Contingency Tables With \pkg{vcd}

### 2005-10-21 trunk/vcd/paper/JSS.Rnw
further fixes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@479 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@

%\VignetteIndexEntry{The Strucplot Framework---Visualizing Multi-way Contingency Tables}
%\VignetteDepends{vcd}
%\VignetteKeywords{graphics, categorical data}
%\VignettePackage{vcd}

@@

  Wirtschaftsuniversit\"at Wien\\
  1090 Wien, Austria\\
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/1/31336-5053
%% Fax: +43/1/31336-734

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, gp = shading_max, split_vertical = TRUE)
@
\caption{Mosaic plot for the arthritis data.}
\label{fig:arthritis}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
cotabplot(UCBAdmissions, panel = cotab_assoc)
@
\caption{Conditional association plot for the UCB admissions data.}
\label{fig:ucbadmissions}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<PreSex,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(PreSex, condvars = c(1, 4), shade = TRUE)
@
\caption{Mosaic plot for the pre- and extramarital sex data.}
\label{fig:presex}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[p]
\begin{center}
<<Titanic,fig=TRUE,echo=FALSE,height=6,width=8,results=hide>>=
doubledecker(Survived ~ ., data = Titanic, labeling_args = 
             list(set_varnames = c(Sex = "Gender")))
@
\caption{Double-decker plot for the survival on the Titanic data.}
\label{fig:titanic}
\end{center}
\end{figure}

@@

\section[The Strucplot Framework]{The Strucplot Framework}
\label{sec:strucplot}

The strucplot framework in the {\sf R} package \pkg{vcd}, used for visualizing
multi-way contingency tables, integrates techniques such as
mosaic displays, association plots, and sieve plots. The main idea is to visualize
the tables' cells arranged in rectangular form. For multi-way tables,
the variables are nested into rows and columns using recursive
conditional splits, given the margins. The result is a 
`flat' representation that can be visualized in
ways similar to a two-dimensional table.
This principle defines a class of conditional displays which allows
for granular control of graphical appearance aspects, including:

@@

\subsection{Mosaic, Association, and Sieve Plots}

As an example, consider the \data{HairEyeColor} 
data containing two polytomous variables (hair and eye color), 
as well as one (artificial) dichotomous variable (sex, i.e., gender). The
`flattened' contingency table can be obtained using the
\codefun{structable} function (quite similar to \codefun{ftable} in
base {\sf R}, but allowing the specification of split directions):

@@

\noindent The small bullets indicate zero entries in the corresponding
cell. Note that in contrast to, e.g.,  \codefun{mosaicplot} in base
{\sf R}, the default split direction and level ordering in all strucplot
displays correspond to the textual representation.
It is also possible to visualize the expected values instead of the
observed values (see Figure \ref{fig:expected}):

@@

\noindent Alternatively, we can directly inspect the residuals. 
The Pearson residuals (standardized deviations of observed from expected values) 
are preferably visualized using association
plots \citep{vcd:Cohen:1980}. In contrast to \codefun{assocplot} in
base {\sf R}, \pkg{vcd}'s \codefun{assoc}
function scales to more than two variables (see Figure \ref{fig:residuals}):

@@

\noindent For compatibility with \codefun{mosaicplot} in base {\sf R},
the \codefun{mosaic} function also allows the use of a \code{"direction"} 
argument taking a vector of \code{"h"} and
\code{"v"} characters (see Figure \ref{fig:split}):

@@

\section{Residual-based Shadings}
\label{sec:shading}

Unlike other graphics functions in base {\sf R}, the strucplot framework 
allows almost full control over the graphical parameters of all plot elements. In
particular, in association plots, mosaic plots, and sieve plots, 
the user can modify the graphical appearance of each tile individually. 
Built on top of this functionality, the
framework supplies a set of shading functions choosing colors
appropriate for the visualization of log-linear models.
The tiles' graphical parameters are set using the \code{gp} argument
of the functions of the strucplot framework. This argument basically expects an object
of class \class{gpar} whose components are arrays of the same shape 
(length and dimensionality) as the data table (see Section \ref{sec:gp}). 
For convenience, however, the user can also
supply a specialized graphical appearance control (``grapcon'') 
function that computes such an object given a vector of
residuals, or, alternatively, a generating function that takes
certain arguments and returns such a grapcon function (see Section
\ref{sec:shadingcustom}). We provide several shading functions, including
support for both HSV and HCL colors, and the
visualization of significance tests (see Section \ref{sec:overview}).

@@

We simply wrap our grapcon shading function in another 
function that takes all additional arguments it needs to use, possibly
preprocesses them, and returns the actual shading function. This returned
function will have access to the parameters since in {\sf R}, nested
functions are lexically scoped. Thus, the grapcon generator
returns (`creates') a `parameterized' shading function with the minimal standard interface
\codefun{strucplot} requires. The following example shows the necessary
extensions for our running example:

@@

\noindent (The \code{zero\_size = 0} argument removes the bullets indicating zero 
observed values. This feature is not provided
in the original {\sf SAS} implementation of the Friendly mosaic plots.)

@@

\section[Flexible Labeling]{Flexible Labeling}
\label{sec:labeling}

One of the major enhancements in package \pkg{vcd} compared to
\codefun{mosaicplot} and \codefun{assocplot} in
base {\sf R} is the labeling in the strucplot framework which 
offers many more features and flexibility. 
Like shading, spacing, and drawing of legend and core plot, 
labeling is now carried out by grapcon functions, rendering labeling completely
modular.
The user supplies either a labeling function, or, alternatively, 
a generating function that parameterizes a labeling function, 
to \codefun{strucplot} which then draws the labels.
Labeling is well-separated from the actual plotting that occurs in the
low-level core functions. It only relies on the viewport tree produced by them, and the
`\texttt{dimnames}' attribute of the visualized table. Labeling
functions are grapcons that ``add ink to the canvas'': the drawing of the
labels happens after the actual plot has been drawn by the core function.
Thus, it is possible to supply one's own labeling function, or
to combine some of the basic functions to produce a more complex labeling.
In the following, we describe the three basic modules
(\codefun{labeling\_text}, \codefun{labeling\_list}, and \codefun{labeling\_cells}) and
derived functions that build upon them.

@@

\codefun{labeling\_text} is the default for all strucplot displays. It
plots labels in the borders similar to the \codefun{mosaicplot}
function in base {\sf R}, but is much more flexible: it is not limited
to 4 dimensions, and the positioning and graphical parameters 
of levels and variable names are customizable. In addition,
the problem of overlapping labels can be handled in several ways.

@@

\noindent The \code{abbreviate} argument takes a vector of
integers indicating the number of significant characters the levels should be
abbreviated to (\code{TRUE} is interpreted as 1, obviously). Abbreviation
is performed using the \codefun{abbreviate} function in base {\sf R}. Another
possibility is to rotate the levels (see Figure~\ref{fig:rotating}):

@@

\section{Customizable Spacing}
\label{sec:spacing}

Spacing of strucplot displays is customizable in a similar way than shading. The
\code{spacing} argument of the \codefun{strucplot} function takes a
list of \class{unit} vectors, one for each dimension, specifying the
space between the tiles corresponding to the levels. As an example,
consider the introductory example of the \data{arthritis} data 
(Figure \ref{fig:arthritis}). Since we are interested in the effect of
the medicament in the placebo and treatment groups, a mosaic plot is
certainly appropriate to visualize the three levels of `Improved' in
the two `Treatment' strata. However, following a slightly different approach,
we might as well want to \emph{highlight} patients with `Marked' improvement in both
groups, that is, produce a spine plot. To obtain such a display within 
the strucplot framework\footnote{The \pkg{vcd} package also provides 
the \codefun{spine} function for spine plots and spinograms.}, it
suffices to set the space between the `Improved' tiles to 0 (see
Figure \ref{fig:artspine}):

@@

\begin{figure}[p]
\begin{center}
<<artspinefig,echo=FALSE,fig=TRUE>>=
<<artspine>>
@ 
\caption{Spine plot for the \data{arthritis} data using the strucplot framework.}
\label{fig:artspine}
\end{center}
\end{figure}

@@

>>>>

\usepackage{thumbpdf}

@@

%\VignetteIndexEntry{The Strucplot Framework: Visualizing Multi-way Contingency Tables With package vcd}
%\VignetteDepends{vcd}
%\VignetteKeywords{contingency tables, mosaic plots, association plots, sieve plots, R, categorical data, independence, conditional independence, HSV, HCL, residual-based shading, grid}
%\VignettePackage{vcd}

@@

  Wirtschaftsuniversit\"at Wien\\
  Augasse 2--6
  1090 Wien, Austria\\
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/1/31336-5053
%% Fax: +43/1/31336-734

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, gp = shading_max, split_vertical = TRUE)
@
\caption{Mosaic plot for the \data{Arthritis} data.}
\label{fig:arthritis}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
cotabplot(UCBAdmissions, panel = cotab_assoc)
@
\caption{Conditional association plot for the \data{UCBAdmissions} data.}
\label{fig:ucbadmissions}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<PreSex,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(PreSex, condvars = c(1, 4), shade = TRUE)
@
\caption{Mosaic plot for the \data{PreSex} data.}
\label{fig:presex}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[p]
\begin{center}
<<Titanic,fig=TRUE,echo=FALSE,height=6,width=8,results=hide>>=
doubledecker(Survived ~ ., data = Titanic, labeling_args = 
             list(set_varnames = c(Sex = "Gender")))
@
\caption{Double-decker plot for the \data{Titanic} data.}
\label{fig:titanic}
\end{center}
\end{figure}

@@

\section[The strucplot framework]{The strucplot framework}
\label{sec:strucplot}

The strucplot framework in the \proglang{R} package \pkg{vcd}, used for visualizing
multi-way contingency tables, integrates techniques such as
mosaic displays, association plots, and sieve plots. The main idea is to visualize
the tables' cells arranged in rectangular form. For multi-way tables,
the variables are nested into rows and columns using recursive
conditional splits, given the margins. The result is a 
`flat' representation that can be visualized in
ways similar to a two-dimensional table.
This principle defines a class of conditional displays which allows
for granular control of graphical appearance aspects, including:

@@

\subsection{Mosaic, association, and sieve plots}

As an example, consider the \data{HairEyeColor} 
data containing two polytomous variables (hair and eye color), 
as well as one (artificial) dichotomous variable (sex, i.e., gender). The
`flattened' contingency table can be obtained using the
\codefun{structable} function (quite similar to \codefun{ftable} in
base \proglang{R}, but allowing the specification of split directions):

@@

\noindent The small bullets indicate zero entries in the corresponding
cell. Note that in contrast to, e.g.,  \codefun{mosaicplot} in base
\proglang{R}, the default split direction and level ordering in all strucplot
displays correspond to the textual representation.
It is also possible to visualize the expected values instead of the
observed values (see Figure \ref{fig:expected}):

@@

\noindent Alternatively, we can directly inspect the residuals. 
The Pearson residuals (standardized deviations of observed from expected values) 
are preferably visualized using association
plots \citep{vcd:Cohen:1980}. In contrast to \codefun{assocplot} in
base \proglang{R}, \pkg{vcd}'s \codefun{assoc}
function scales to more than two variables (see Figure \ref{fig:residuals}):

@@

\noindent For compatibility with \codefun{mosaicplot} in base \proglang{R},
the \codefun{mosaic} function also allows the use of a \code{"direction"} 
argument taking a vector of \code{"h"} and
\code{"v"} characters (see Figure \ref{fig:split}):

@@

\section{Residual-based shadings}
\label{sec:shading}

Unlike other graphics functions in base \proglang{R}, the strucplot framework 
allows almost full control over the graphical parameters of all plot elements. In
particular, in association plots, mosaic plots, and sieve plots, 
the user can modify the graphical appearance of each tile individually. 
Built on top of this functionality, the
framework supplies a set of shading functions choosing colors
appropriate for the visualization of log-linear models.
The tiles' graphical parameters are set using the \code{gp} argument
of the functions of the strucplot framework. This argument basically expects an object
of class \class{gpar} whose components are arrays of the same shape 
(length and dimensionality) as the data table (see Section \ref{sec:gp}). 
For convenience, however, the user can also
supply a specialized graphical appearance control (``grapcon'') 
function that computes such an object given a vector of
residuals, or, alternatively, a generating function that takes
certain arguments and returns such a grapcon function (see Section
\ref{sec:shadingcustom}). We provide several shading functions, including
support for both HSV and HCL colors, and the
visualization of significance tests (see Section \ref{sec:overview}).

@@

We simply wrap our grapcon shading function in another 
function that takes all additional arguments it needs to use, possibly
preprocesses them, and returns the actual shading function. This returned
function will have access to the parameters since in \proglang{R}, nested
functions are lexically scoped. Thus, the grapcon generator
returns (`creates') a `parameterized' shading function with the minimal standard interface
\codefun{strucplot} requires. The following example shows the necessary
extensions for our running example:

@@

\noindent (The \code{zero\_size = 0} argument removes the bullets indicating zero 
observed values. This feature is not provided
in the original \proglang{SAS} implementation of the Friendly mosaic plots.)

@@

\section[Flexible labeling]{Flexible labeling}
\label{sec:labeling}

One of the major enhancements in package \pkg{vcd} compared to
\codefun{mosaicplot} and \codefun{assocplot} in
base \proglang{R} is the labeling in the strucplot framework which 
offers many more features and flexibility. 
Like shading, spacing, and drawing of legend and core plot, 
labeling is now carried out by grapcon functions, rendering labeling completely
modular.
The user supplies either a labeling function, or, alternatively, 
a generating function that parameterizes a labeling function, 
to \codefun{strucplot} which then draws the labels.
Labeling is well-separated from the actual plotting that occurs in the
low-level core functions. It only relies on the viewport tree produced by them, and the
`\texttt{dimnames}' attribute of the visualized table. Labeling
functions are grapcons that ``add ink to the canvas'': the drawing of the
labels happens after the actual plot has been drawn by the core function.
Thus, it is possible to supply one's own labeling function, or
to combine some of the basic functions to produce a more complex labeling.
In the following, we describe the three basic modules
(\codefun{labeling\_text}, \codefun{labeling\_list}, and \codefun{labeling\_cells}) and
derived functions that build upon them.

@@

\codefun{labeling\_text} is the default for all strucplot displays. It
plots labels in the borders similar to the \codefun{mosaicplot}
function in base \proglang{R}, but is much more flexible: it is not limited
to 4 dimensions, and the positioning and graphical parameters 
of levels and variable names are customizable. In addition,
the problem of overlapping labels can be handled in several ways.

@@

\noindent The \code{abbreviate} argument takes a vector of
integers indicating the number of significant characters the levels should be
abbreviated to (\code{TRUE} is interpreted as 1, obviously). Abbreviation
is performed using the \codefun{abbreviate} function in base \proglang{R}. Another
possibility is to rotate the levels (see Figure~\ref{fig:rotating}):

@@

\section{Customizable spacing}
\label{sec:spacing}

Spacing of strucplot displays is customizable in a similar way than shading. The
\code{spacing} argument of the \codefun{strucplot} function takes a
list of \class{unit} vectors, one for each dimension, specifying the
space between the tiles corresponding to the levels. As an example,
consider the introductory example of the \data{Arthritis} data 
(Figure \ref{fig:arthritis}). Since we are interested in the effect of
the medicament in the placebo and treatment groups, a mosaic plot is
certainly appropriate to visualize the three levels of `Improved' in
the two `Treatment' strata. However, following a slightly different approach,
we might as well want to \emph{highlight} patients with `Marked' improvement in both
groups, that is, produce a spine plot. To obtain such a display within 
the strucplot framework\footnote{The \pkg{vcd} package also provides 
the \codefun{spine} function for spine plots and spinograms.}, it
suffices to set the space between the `Improved' tiles to 0 (see
Figure \ref{fig:artspine}):

@@

\begin{figure}[p]
\begin{center}
<<artspinefig,echo=FALSE,fig=TRUE>>=
<<artspine>>
@ 
\caption{Spine plot for the \data{Arthritis} data using the strucplot framework.}
\label{fig:artspine}
\end{center}
\end{figure}

@@

\begin{appendix}
\section{Data sets}

The data set names in the paper are those from the \proglang{R}
system. In the following, we give a short description of each data set.

\begin{description}
\item[Titanic] Information on the fate of passengers on
  the fatal maiden voyage of the ocean liner `Titanic', summarized
  according to economic status (class), gender (`Sex'), age and
  survival. The data were originally collected by the British Board of Trade
  in their investigation of the sinking. Source: FIXME. Available in base \proglang{R}.
\item[PreSex]  Data from Thornes \& Collard (1979), reported in Gilbert (1981), on
  pre- and extra-marital sex and divorce. Source: FIXME. Available
  in the \pkg{vcd} package.
\item[Arthritis] Data from \cite{vcd:Koch+Edwards:1988} from a double-blind clinical trial
  investigating a new treatment for rheumatoid arthritis. Available
  in the \pkg{vcd} package.
\item[UCBAdmissions] Aggregate data on applicants to graduate school at Berkeley for
  the six largest departments in 1973 classified by admission and
  gender. Source: \cite{vcd:Bickel+Hammel+O'Connell:1975}. Available
  in base \proglang{R}.
\item[HairEyeColor] Distribution of hair and eye color and gender in 592 statistics
  students. The gender information is artificial. Source: \cite{vcd:Snee:1974}.
  Available in base \proglang{R}.
\item[Bundesliga] Results from the first German soccer league in the
  years 1995/6 \citep{vcd:Knorr-Held:1999} and 2001/2 (Achim
  Zeileis). Available in the \pkg{vcd} package.
\end{description}

\end{appendix}
